Logsums

### 2020-05-11 04_accessibility.Rmd
Restructure accessibility calculation
This makes better use of the tidyverse, simplifying the logsum code. Also, discusses Euclidean vs network distance in more detail.

---
title: "Logsums"
author: "Greg Macfarlane"
date: "6/7/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(sf)
library(leaflet)
library(lubridate)
library(VGAM)
```

In this document we calculate the utility-based measure of accessibility, as well
as the 10-minute walk buffer.

## Distances
We cleaned and gathered the data parks data and data for tracts in the previous
two pages, as well as computed the network-based distance between each block
group and the nearest edge point of each park.

```{r load, message=FALSE}
open_spaces <- read_rds("data/open_spaces.rds")
blockgroups <- read_csv("data/blockgroups.csv")

# tigris not working, manual download
#bg_poly <- tigris::block_groups("New York", county =  c("081", "047", "061", "005", "085"))
bg_poly <- st_read("/Users/gregmacfarlane/Library/Caches/tigris/tl_2019_36_bg.shp") %>%
  transmute(geoid = GEOID, p_water = AWATER / (AWATER + ALAND)) %>%
  filter(p_water < 1) %>%
  filter(geoid %in% blockgroups$GEOID) %>%
  st_transform(32618)
```

In calculating the distances, two things could happen. First, it is possible that
no path is found between the park and the block group centroid; this could be the
result of network miscodings or other errors in the network algorithm. We'll have
to take the results as they come, and we simply remove the park from that block
group's accessibility calculation. The other issue is when a block group centroid
and the park edge point match to the *same* node, resulting in a distance of zero.
Because we are going to log-transform the distance, we add ten meters to all paths.

```{r distance_adjust}
distance_files <- list.files("data/nyc_shortest_paths_3500m/", full.names = TRUE)

read_all_files <- function(pattern = pattern, path = "./"){
  map_df(list.files(path = path, pattern = pattern, full.names = TRUE), read_csv,
         col_types = list(geoid = col_character(), park_id = col_character()))
}
distances <- read_all_files(".csv", "data/nyc_shortest_paths_5000m/") %>%
  # filter out infinite paths
  filter(!is.infinite(distance)) %>%
  # add ten meters to all paths
  mutate(distance = distance + 10)
```

Now we can calculate the utility specification. The first specification uses the
distance and size terms from the Alameda County study. The second specification
adds the amenities with the utilities found by the Kinnell et al. study.

```{r logsums}
logsums <- distances %>%
  left_join(open_spaces %>% st_set_geometry(NULL), by = c("park_id" = "id")) %>%
  mutate(
    distance_miles = (distance + 10) * 0.000621371,
    u_access = -1.768 * log(distance_miles) + 0.373 * log(as.numeric(acres)),
    u_multi = u_access + 
      open_spaces$trails * 0.99 + 
      open_spaces$playgrounds * 0.26 + 
      open_spaces$courts * 0.43
  ) %>%
  group_by(geoid) %>%
  summarise(
    access_ls = log(sum(exp(u_access), na.rm = TRUE)),
    multi_ls = log(sum(exp(u_multi), na.rm = TRUE))
  ) %>%
  mutate_at(
    vars(ends_with("_ls")),
    list(~ (. - mean(., na.rm = TRUE)) / sd(., na.rm = TRUE))
  )

bg_access <- bg_poly %>%
  left_join(logsums, by = "geoid") %>%
  write_rds("data/bg_access.rds")
```

## 10-minute buffer

We are also interested in looking at what happens when the tracts are within a 
10-minute walk of a park, the threshold established by the Trust for Public Lands.
This is interesting because whereas a limited number of tract centroids failed
to overlap with the 10-minute buffer, all of the block group centroids intersect.

## Accessibility Map
Here's a map of the values. Note that there is also a shiny app in this project
that allows the user to set different values for the logsum parameters.

```{r access_map}
# this interactive map is much easier to create and explore
pal <- colorQuantile("GnBu", c(bg_access$access_ls, bg_access$multi_ls), n = 5)
leaflet(bg_access %>% st_transform(4326)) %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addPolygons(group = "access", color = ~ pal(access_ls), stroke = 1,
              label = ~as.character(round(access_ls, 2))) %>%
  addPolygons(group = "Kinner", color = ~ pal(multi_ls), stroke = 1,
              label = ~as.character(round(multi_ls, 2))) %>%
  addLayersControl(overlayGroups = c("access", "Kinner"))
```

>>>>

---
title: "Logsums"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
library(tidyverse)
library(sf)
library(spdep)
library(spatialreg)
library(leaflet)
library(lubridate)
library(VGAM)
```

In this document we calculate the accessiblity logsums based on the Euclidean
and network distances from the tract and block group centroids to the tract,
combined with park attributes.

```{r load}
tracts <- read_rds("data/tracts.rds")
open_spaces <- read_rds("data/open_spaces.rds")
```

## Distance Examination
The distances were computed according to the code presented in the [Distances](03_distances.html)
page. But we want to evaluate the difference between the various distance measurements
and visualize what they mean in terms of our analysis.

```{r read_distances}
blockgroup_files <- list.files("data/blockgroups", "*.csv", full.names = TRUE)
tract_files <- list.files("data/tracts", "*.csv", full.names = TRUE)

read_dist_csv <- function(x){
  # warnings are from `inf` strings, which become converted to NA, which is okay.
  read_csv(x, col_types = "ccnn") %>% as_tibble()
}

blockgroup_distances <- lapply(blockgroup_files, read_dist_csv) %>% bind_rows()
tract_distances      <- lapply(tract_files,      read_dist_csv) %>% bind_rows()
```

First, let's compare the difference between the Euclidean and network
distance at the block group level. Well, this is troubling; it appears as though
there is no relationship between the two methods of measurement.

```{r bg_dist_comparison}
ggplot(blockgroup_distances %>% sample_n(1e4), aes(x = euc_dist, y = distance)) +
  geom_point(alpha = 0.3) + geom_smooth()
```

Let's look at the same plot for tracts. This is much better, though there
are still some curious issues.

```{r tr_dist_comparison}
ggplot(tract_distances %>% sample_n(1e4), aes(x = euc_dist, y = distance)) +
  geom_point(alpha = 0.3) + geom_smooth()
```

In particular, how is it possible for the
Euclidean distance to be *longer* than the network distance? Let's look at this
by distance. It seems that everything is doing okay except for tracts and parks in Staten 
Island (prefix '085' and "R"). Obviously the shortest path isn't handling Staten
Island very well. The Verrazzano-Narrows Bridge is in the network, but I-78 and
the Bayonne Bridge route to Manhattan through New Jersey is not in the network.

```{r tr_dist_comparison_borough}
ggplot(tract_distances %>% sample_n(1e4) %>% 
         mutate(borough = substr(geoid, 3, 5),
                park_borough = substr(park_id, 0, 1),
                park_borough = ifelse(!park_borough %in% c("B", "M", "Q", "R", "X"), "C", park_borough)), 
       aes(x = euc_dist, y = distance)) +
  geom_abline(slope = 1, intercept = 0, lty = "dotted", color = "grey") + 
  geom_point(alpha = 0.3) + geom_smooth() +
  xlab("Euclidean Distance") + ylab("Network Distance") +
  facet_grid(borough ~ park_borough) +
  theme_bw()
```

All of this is to say we probably need to eliminate Staten Island for our 
analysis, or else adjust the network distances we calculate.

Is there a relation between the tract distance and the distances for block
groups within the tract?

```{r comparison}
comparison <- blockgroup_distances %>%
  sample_n(1e5) %>%
  rename(bg_net = distance, bg_euc = euc_dist) %>%
  mutate(tract = substr(geoid, 1, 11)) %>%
  left_join(tract_distances %>% rename(tr_net = distance, tr_euc = euc_dist), 
            by = c("tract" = "geoid", "park_id" = "park_id"))
```

Let's look first at the network distances. Yikes!
```{r comparison-net}
ggplot(comparison, aes(x = bg_net, y = tr_net)) +
  geom_point(alpha = 0.1) + geom_smooth() + 
  xlab("Blockgroup Network Distance") + ylab("Tract Network Distance")
```

The Euclidean distances make *loads* more sense, with some aggregation error
but it's obvious the numbers aren't stupid.

```{r comparison-net}
ggplot(comparison, aes(x = bg_euc, y = tr_euc)) +
  geom_point(alpha = 0.1) + geom_smooth() + 
  xlab("Blockgroup Euclidean Distance") + ylab("Tract Euclidean Distance")
```

Let's make a map of tracts colored by how far the two methods think they are
from Central Park. In general this seems fine, except again, Staten Island does
not look that great. Something is definitely happening there.

```{r distance-map}
pal <- colorQuantile(palette = "magma", log(tract_distances$distance), n = 7)
tracts %>%
  select(geoid) %>%
  left_join(tract_distances %>% filter(park_id == "M010"), by = "geoid") %>%
  st_transform(4326) %>%
  leaflet() %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addPolygons(color = ~pal(log(distance)), group = "Network") %>%
  addPolygons(color = ~pal(log(euc_dist)), group = "Euclidean") %>%
  addLayersControl(overlayGroups = c("Network", "Euclidean"))
```

## Logsums

Now that we have the distances between parks and block groups / tracts, we can 
calculate the logsums.

```{r logsums}
logsum <- function(x) { log(sum(exp(x))) }

compute_logsums <- function(df) {
  df %>%
    left_join(open_spaces %>% st_set_geometry(NULL), by = c("park_id" = "id")) %>%
    # use euc_distance for impossible distances.
    mutate( distance = ifelse(distance < euc_dist, euc_dist, distance) ) %>%
    # convert distances to miles
    mutate_at(c("distance", "euc_dist"), ~ . * 0.000621371) %>%
    # must be at least 1/10 miles
    mutate_at(c("distance", "euc_dist"), ~ ifelse(. < 0.1, 0.1, .)) %>%

    # compute utility
    mutate(
      euc_access = -1.768 * log(euc_dist) + 0.373 * log(as.numeric(acres)),
      euc_multi  = euc_access + trails * 0.99 +  playgrounds * 0.26 + courts * 0.43,
      net_access = -1.768 * log(distance) + 0.373 * log(as.numeric(acres)),
      net_multi  = net_access + trails * 0.99 +  playgrounds * 0.26 + courts * 0.43
    ) %>%

    # group by origin and compute access logsum
    group_by(geoid) %>%
    summarize_at(vars(euc_access:net_multi), logsum) %>%

    # standardize
    mutate_at(vars(euc_access:net_multi), ~(. - mean(.)) / sd(.))

}

tract_logsums <- compute_logsums(tract_distances) %>% write_rds("data/tract_logsums.rds")
bg_logsums    <- compute_logsums(blockgroup_distances) %>% write_rds("data/bg_logsums.rds")
```

Let's visualize the access logsums calculated for each graph in this method.

```{r tracts_access}
tracts_access <- tracts %>% left_join(tract_logsums, by = "geoid")
# this interactive map is much easier to create and explore
pal <- colorQuantile("GnBu", c(tracts_access$euc_access, tracts_access$net_multi), n = 7)
leaflet(tracts_access %>% st_transform(4326)) %>%
  addProviderTiles(providers$Esri.WorldGrayCanvas) %>%
  addPolygons(group = "Network", color = ~ pal(net_access),
              label = ~as.character(round(net_access, 2))) %>%
  addPolygons(group = "Network-Multi", color = ~ pal(net_multi),
              label = ~as.character(round(net_multi, 2))) %>%
  addPolygons(group = "Euclidean", color = ~ pal(euc_access),
              label = ~as.character(round(euc_access, 2))) %>%
  addPolygons(group = "Euclidean-Multi", color = ~ pal(euc_multi),
              label = ~as.character(round(euc_multi, 2))) %>%
  addLayersControl(baseGroups = c("Network", "Euclidean", "Network-Multi", "Euclidean-Multi"))
```
